### 2019-03-21 paper/sex_selection_spacing-ver2.tex
Expand on distribution results section and update distribution part of Introduction

% [Hypotheses / questions]
Despite the increasing use of sex selection, the important impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis I examine here is that spacing patterns significantly changed 
as prenatal sex determination became available.
I address three questions in order of importance.

First, how did birth intervals change over time for different groups and how are those 
changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, whether urban or rural, and the sex composition of 
prior children.

[TK change based on rewrite of section]
Second, have the shapes of the survival curves changed?
The shapes of the survival curves provide information both on general changes in birth 
spacing and whether parents are still trying to conceive earlier after girls than boys.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the duration between births, 
and may even lead to longer average spacing after the birth of girls than after boys.
I expect this to be the case if there is still a strong preference for boys, and this 
behavior may even become more pronounced with access to sex selection.

@@

Average spacing is a convenient way to understand the overall changes but may hide 
important differences in the distribution of spacing.
Specifically, a concern is that the increases in average spacing that I find above may 
mask an increased variance in birth spacing for families with fewer sons if they are both 
more likely to have a very early pregnancy and to continue through pregnancies and 
sex-selective abortions until they conceive a son.
This section, therefore, provides more detail on how spacing is distributed within a spell 
and across sex compositions using a graphical approach.

I show survival curves conditional on predicted parity progression rather than standard 
survival curves. 
The advantage of this approach is that it is possible to directly compare the distribution 
of spacing to next birth across groups, independently of differences in how likely the 
next birth is.
Because the conditional survival curves are independent of the likelihood of parity 
progression, they all begin at 100\% and end at 0\%.

Instead of averaging across the entire sample, I calculate the conditional survival curves 
for an average woman using the method detailed in Section \ref{sec:strategy}.%
\footnote{
To complement this approach, Appendix Tables \ref{tab:p25_p50_p75_low}, 
\ref{tab:p25_p50_p75_med}, and \ref{tab:p25_p50_p75_high} show 25th, 50th, and 75th 
percentiles durations with bootstrapped standard errors based on averages across all women 
in the each subsample as for the results above.
} 
For each combination of education and spell, I use values based on the average age at the 
start of the spell.
Furthermore, I use the majority categories for the categorical explanatory  variables, 
which means no ownership of land and not in a scheduled caste or tribe.
The characteristics used do not change across the four decades to ensure that composition 
effects do not drive any changes.

In the interest of brevity, I discuss only a select set of subsamples.
Figures \ref{fig:pps_low} and \ref{fig:pps_high} show spacing across sex compositions for 
the second, third, and fourth spells for the first period, 1972--1984, and the last 
period, 2005--2016, for rural women with no education and for urban women with eight or 
more years of education.
The Appendix shows conditional survival curves for all groups and decades.

@@

Figure \ref{fig:pps_low} shows the distribution of birth within a spell for women with no 
education.
For the second spell, what is most interesting is that there is almost no difference 
between those with a first-born boy and those with a first-born girl.
For the 1972--1984 period the two conditional survival curves only begin to diverge 
at around 18 months (corresponding to 27 months after the first birth took place), and
at that point close to 50 percent of the women have already had their second child.
The literature on birth spacing and child health suggests that an interpregnancy interval
of less than 18 months is associated with a higher risk of adverse health outcomes and
mortality \citep{Conde-Agudelo2012}.
With a nine-month pregnancy this means that almost 50 percent of the children born fall
in this category, and, somewhat surprisingly, that there do not appear to be any particular 
disadvantage for the child born after a girl than after a boy.%
\footnote{
Note that this does not imply that the first-born girl is not in a more adverse position
than a first-born boy when the second child arrives, just that this is not directly
related to a short birth interval.
}
The difference in spacing between a first-born girl and a first-born boy first-born 
remains almost the same across the four decades.
Furthermore, the proportion with very short spacing remains almost the same.

@@

The correlation between the increased use of sex selection and the reversal of the 
standard spacing pattern show up clearly in the 2005--2016 period for the third spell with 
spacing to the third child consistently longer with two girls than with either two boys 
or one boy and one girl throughout the entire spell.%
\footnote{
For the third spell, the 1972--1984 period show a close to standard pattern throughout, 
although, as mentioned above, the sex ratios for women without girls was substantially 
higher than the natural sex ratio either because of recall error or early use of 
sex selection.
}
There is an even more pronounced rightward and upward shift for spacing after only
girls in the fourth spell for the first years of the spell.%
\footnote{
The caveat is that there are few births to base the results on,
especially for women with two or more sons.
} 
Furthermore, the differences are substantial across most of the distribution.
Twelve months after the beginning of the spell there is a more than ten 
percentage points difference in the conditional survival curves for only girls
and only boys.
Both of the spells therefore show changes in the distribution of births that is in line
with extensive use of sex selection among families with no or one son.

% Even with sex selection, however, the observed pattern for the third and fourth spells
% is not preordained.
% We should expect 51.2 percent of the women to conceive a boy when they have their first 
% pregnancy in the spell, which means that they do not need to use sex selection to have
% a son.
% Hence, even with sex selection, we might expect to observe a very short interval for the 
% initial births if families with strong son preference still try to conceive sooner than 
% those with lower son preference.
% That we do not observe a short intervals for the early births calls into question the 
% assumption that son preference leads to very early births with fewer sons even in 
% situations when there is access to sex selection.
% If, however, parents know that children born after short interpregnancy periods suffer 
% from worse health outcomes it makes sense to wait longer before conceiving again in
% this situation, precisely because they know that with sex selection the next child they 
% carry to term will be a boy.
% 
% Hence, there are two potential reasons why son preference combined with access to
% sex selection can lead to longer birth intervals.
% First, the ``abortion'' effect, where each abortion extend birth spacing by between six
% and twelve months.
% Second, a ``delay'' effect, where parents conceive later because they know that short
% spacing is detrimental to the next child's health, and that the next child is more likely
% to be a boy because of access to sex selection.
% The ``delay'' effect may explain why the twenty-five percentile birth intervals for the
% second spell increase more for families with two girls than either of the other two
% sex compositions, even though there should be few if any sex-selective abortions among
% these births.%
% \footnote{
% See Appendix Table \ref{tab:p25_p50_p75_high}.
% }

>>>>

% [Hypotheses / questions]
Despite the increasing use of sex selection, the important impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis is that spacing patterns significantly changed as prenatal sex 
determination became available.
Specifically, I address three questions.

First, how did average birth intervals change over time for different groups and how are 
those changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
I still expect parents to \emph{try} to conceive earlier after the birth of a girl than 
after a boy, but the abortions will increase the duration between births, and may even 
lead to longer average spacing after the birth of girls than after boys.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, whether urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Very short birth intervals are associated with worse health outcomes and increasing 
mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until a son is conceive.

@@

The results above show consistent increases in average spacing across the board.
Average spacing is a convenient way to understand the overall changes in behavior but may 
hide important differences in the distribution of spacing.
Specifically, whether these increases in average spacing come from fewer births after 
very short intervals, because of even longer spacing from births with already long 
spacing, or a general increase in spacing have important implications for what effects the
increase in average spacing are likely to have and how to interpret the motivation for
the changes.
On one hand, if there still are many births that occur after very short intervals, the 
increase in average spacing is unlikely to have a substantial impact on health outcomes
since the literature on birth spacing and child health suggests that an interpregnancy 
interval of less than 18 months is associated with a higher risk of adverse health 
outcomes and mortality \citep{Conde-Agudelo2012}.
On the other hand, we may see a reduction in early observed births and more long spells if 
families with fewer sons are both more likely to have a very early pregnancy and to 
continue through pregnancies and sex-selective abortions until they conceive a son.%
\footnote{
It is unclear what the optimal response in timing of pregnancies to the availability of 
sex selection should be.
The expected birth intervals become longer with the use of sex selection, which may be 
costly for parents leading them to try to conceive even earlier than previously.
Conversely, if parents know that short birth intervals are detrimental to the next child's
health, and with sex selection the next child is likely be a boy, it might make sense to 
wait longer before conceiving.
}

This section, therefore, provides more detail on how spacing is distributed within a spell 
and across sex compositions using a graphical approach.
I show survival curves conditional on predicted parity progression rather than standard 
survival curves. 
The advantage of this approach is that it is possible to directly compare the distribution 
of spacing to next birth across groups, independently of differences in how likely the 
next birth is.
Because the conditional survival curves are independent of the likelihood of parity 
progression, they all begin at 100\% and end at 0\%.

Instead of averaging across the entire sample, I calculate the conditional survival curves 
for an average woman using the method detailed in Section \ref{sec:strategy}.
For each combination of education and spell, I use values based on the average age at the 
start of the spell.
Furthermore, I use the majority categories for the categorical explanatory  variables, 
which means no ownership of land and not in a scheduled caste or tribe.
The characteristics used do not change across the four decades to ensure that composition 
effects do not drive any changes.
To complement this approach, Appendix Tables \ref{tab:p25_p50_p75_low}, 
\ref{tab:p25_p50_p75_med}, and \ref{tab:p25_p50_p75_high} show 25th, 50th, and 75th 
percentiles durations with bootstrapped standard errors based on averages across all women 
in the each subsample as for the results above.

In the interest of brevity, I discuss only two subsamples.
Figures \ref{fig:pps_low} and \ref{fig:pps_high} show spacing across sex compositions for 
the second, third, and fourth spells for the first period, 1972--1984, and the last 
period, 2005--2016, for rural women with no education and for urban women with eight or 
more years of education.
The Appendix shows conditional survival curves for all groups and decades.

@@

Figure \ref{fig:pps_low} shows the distribution of birth within a spell for women with no 
education.
For the second spell, what is most interesting is that there is almost no difference 
between those with a first-born boy and those with a first-born girl.
For the 1972--1984 period the two conditional survival curves only begin to diverge 
at around 18 months (corresponding to 27 months after the first birth took place), and
at that point close to 50 percent of the women have already had their second child.
With a nine-month pregnancy, this means that almost 50 percent of the children born are
born within the interpregnancy interval associated with a higher risk of adverse health
outcomes and mortality, and, somewhat surprisingly, that there do not appear to be any 
particular disadvantage for the child born after a girl than after a boy.%
\footnote{
Note that this does not imply that the first-born girl is not in a more adverse position
than a first-born boy when the second child arrives, just that this is not directly
related to a short birth interval.
}
The difference in spacing between a first-born girl and a first-born boy first-born 
remains almost the same across the four decades.
Furthermore, the proportion with very short spacing remains almost the same.

@@

For the third spell, the 1972--1984 period show a close to standard pattern throughout, 
although, as mentioned above, the sex ratios for women without girls was substantially 
higher than the natural sex ratio either because of recall error or early use of 
sex selection.
The correlation between the increased use of sex selection and the reversal of the 
standard spacing pattern show up clearly in the 2005--2016 period for the third spell with 
spacing to the third child consistently longer with two girls than with either two boys 
or one boy and one girl throughout the entire spell.
This pattern is most consistent with early and continued use of sex selection, rather 
than a general reduction in the likelihood of very short birth intervals.
It also means, that previous girls can expect a longer period without a younger sibling,
which may reduce the sibling competition for resources.

There is an even more pronounced rightward and upward shift for spacing after only
girls in the fourth spell for the first years of the spell.%
\footnote{
The caveat is that there are few births to base the results on,
especially for women with two or more sons.
} 
Furthermore, the differences are substantial across most of the distribution.
Twelve months after the beginning of the spell there is a more than ten 
percentage points difference in the conditional survival curves for only girls
and only boys.
Both of the spells therefore show changes in the distribution of births that is in line
with extensive use of sex selection among families with no or one son.

% We should expect 51.2 percent of the women to conceive a boy when they have their first 
% pregnancy in the spell, which means that they do not need to use sex selection to have
% a son.
% Hence, even with sex selection, we might expect to observe a very short interval for the 
% initial births if families with strong son preference still try to conceive sooner than 
% those with lower son preference.
% If, however, parents know that children born after short interpregnancy periods suffer 
% from worse health outcomes it makes sense to wait longer before conceiving again in
% this situation, precisely because they know that with sex selection the next child they 
% carry to term will be a boy.
% 
% Hence, there are two potential reasons why son preference combined with access to
% sex selection can lead to longer birth intervals.
% First, the ``abortion'' effect, where each abortion extend birth spacing by between six
% and twelve months.
% Second, a ``delay'' effect, where parents conceive later because they know that short
% spacing is detrimental to the next child's health, and that the next child is more likely
% to be a boy because of access to sex selection.
